---
title: "Practical Fall 2019"
author: "Kevin Huang"
date: "10/28/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/UMich_Bio201_F19/")
```

As with each lab, add "completed" to the Rmd filename; update the name in the author line above; update the location of the working directory as necessary. 

# Load packages
```{r Load packages, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(readxl)
library(broom)
library(cowplot)
set.seed(7)
```

# Part 1 - concepts 
Type the answers to questions 1-5 in the space below. 

### Question 1

Given vector v1 below, will it be read as character or numeric? Why does this happen? [1 point]
```{r}
v1 <- c("one", "2", "three", "4")
```

The vector v1 will be read as character. This happens because the quotes around the inputs (i.e "one", "2", etc.) tells R they are characters 

### Question 2

dataframe1 has 323 rows and dataframe2 has 345 rows. After an inner join the resulting data fame has 312 rows. Why does the result have fewer rows? [1 point]

The result has fewer rows because inner join excludes groups/samples that exist in one of the data frames, but not in the other. The resulting data frame will thus have fewer rows. 

### Question 3

What are the three assumptions for a t-test? Explain what to do if the variance assumption is violated. Explain what to do if the normality assumption is violated. [3 points]

For a t-test, the first assumption is that the data has equal variance. Second, the data collected is representative, random, and good sample size of the total population. Third, the data has normal (bell-shaped curve) distribution. If the normality assumption is violated, a Wilcoxon test should be used. If the variance assumption is violated, an unpaired t-test should be used.

### Question 4

You have a data frame (called butyrate_df) with the columns: "Participant_ID", "Study_week", "Semester", "Supplement_consumed", "Avg_butyrate". If this data frame is piped to each of the functions below, evaluate if the description of the results is true or false. [3 points]
```{r eval=FALSE, include=FALSE}
butyrate_df %>%
  filter(Semester == "Fall2019")
# Result1 = keeps only data from Fall 2019 semester

butyrate_df %>%
  filter(Avg_butyrate <= 47.0)
# Result2 = keeps only values equal to 47 mmol/kg 

butyrate_df %>%
  group_by(Semester) %>%
  summarize(New_column = median(Avg_butyrate))
# Result3 = new_column contains the median of each participant 
```

For Result1, the result is true. The code will filter only for the data from where the Semester is equal to "Fall2019". For Result2, the result is false. (Avg_butyrate <= 47.0) returns values that are greater than or equal to 47.0, not only 47.0. For Result3, the result is false. New_column will contain the median for each semester, not participant. 

### Question 5

something about NAs

The data frame (called breath_df) has the columns: "Participant_ID", "Study_week", "Semester", "Supplement_consumed", "Methane", "Hydrogen", "Carbon_dioxide". This data frame is used for the code below, explain what will happen to any NAs in each of the shown calculations, and how the NAs will effect the underlying data. [2 points]
```{r eval=FALSE, include=FALSE}
# Option 1
breath_df %>%
  filter(!is.na(Methane)) %>% 
  group_by(Participant_ID, Study_week) %>%
  summarize(Avg_methane = mean(Methane))

# Option 2
breath_df %>%
  group_by(Participant_ID, Study_week) %>%
  summarize(Avg_methane = mean(Methane, na.rm = TRUE))
```

Both options will return the same data. Option 1 filters the data at the beginning to disregard NA values, while Option 2 removes the NA values right before calculating the mean. 

# Part 2 - tidy data 

### Question 6

Find the data frame matching your unique name in the GitHub repository. Find the breath gas and taxa abundance data frames. Download and import all into RStudio and use these data frames to complete the rest of the practical. Name the dataframe to match your uniqname, name the breath gas data frame "breath", name the taxa abundance data frame "taxa". [1 point]
```{r}
# data import code 
kghuang <- read_delim("Raw_data/kghuang.txt", 
    "\t", escape_double = FALSE, trim_ws = TRUE)
breath <- read_delim("Raw_data/breath.txt", 
    "\t", escape_double = FALSE, trim_ws = TRUE)
taxa <- read_delim("Raw_data/taxa.txt", "\t", 
    escape_double = FALSE, trim_ws = TRUE)
```

Update column names in all files match the course conventions. [1 point]
```{r}
kghuang <- kghuang %>%
  rename_all(tolower) 
breath <- breath %>%
  rename_all(tolower)
taxa <- taxa %>%
  rename_all(tolower)

```

View the supplement types in your semester. Are they all entered in the same style? If not update them to match. [1 point]
```{r}
kghuang[["supplement_consumed"]] <- recode(.x=kghuang[["supplement_consumed"]], "Inulin"="inulin")
kghuang[["supplement_consumed"]] <- recode(.x=kghuang[["supplement_consumed"]], "HiMaize"="himaize")
kghuang[["supplement_consumed"]] <- recode(.x=kghuang[["supplement_consumed"]], "Accessible"="accessible")
```

### Question 7

What are the dimensions of each data frame? Write the results as a comment after the function. [1 point]
```{r}
dim(kghuang) # 382 by 9
dim(taxa) # 46881 by 6
dim(breath) # 2668 by 5
```

Select a join that will drop as few data points as possible. Join the data frame containing the participant data with the breath gas data. Name this "joined_df". [1 point]
```{r}
joined_df <- full_join(kghuang, breath)
```

What are the dimensions of the joined data frame? Write the results as a comment after the function. [1 point]
```{r}
dim(joined_df) # 3846 by 11
```

### Question 8

Calculate the total SCFAs. [1 point]
```{r}
joined_df <- joined_df %>%
  mutate(total_scfa = acetate_mmol_kg + butyrate_mmol_kg + propionate_mmol_kg)
```

Calculate the weekly mean of each SCFA of each participant. Calculate the weekly mean methane and hydrogen levels of each participant. [1 point]
```{r}
joined_df <- joined_df %>%
  group_by(participant_id, study_week, semester, supplement_consumed) %>%
  mutate(mean_acetate = mean(acetate_mmol_kg, na.rm = TRUE), 
            mean_butyrate = mean(butyrate_mmol_kg, na.rm = TRUE), 
            mean_propionate = mean(propionate_mmol_kg, na.rm = TRUE))

joined_df <- joined_df %>%
  group_by(participant_id, study_week, semester, supplement_consumed) %>%
  mutate(mean_ch4 = mean(ch4, na.rm = TRUE),
            mean_h2 = mean(h2, na.rm = TRUE))
```

What are the dimensions of the data frame with the avearges? Write the results as a comment after the function.
```{r}
dim(joined_df) # 3846 by 17
```

Join this data frame with the one called "taxa". Name the new data frame "all_data". What are the dimensions of this new data frame? [1 point]
```{r}
all_data <- full_join(joined_df, taxa)
dim(all_data) # 133581 by 21
```

Save all data frames generated to your GitHub page. [1 point]
```{r}
save(all_data, file = "~/all_data.Rda")
save(joined_df, file = "~/joined_df.Rda")
```

# Part 3 - statistics & graphics 

Update all axes labels and figure titles so they are informative and easy to read. Upload final plots to GitHub. When specified, conduct the appropriate tests, and write the statistic (p, F, R value) as a comment. 

### Question 9

If you have more than one type of supplement consumed in your semester, filter for one type to answer Question 9. Plot each SCFA with the study week (weeks 1 and 3 only) on the x-axis. Make each study week a different color. Hint: you may have to use facets, or create individual plots and combine them into a multi-panelled figure. [3 points]
```{r}
plot_ace <- kghuang %>%
  filter(supplement_consumed == "inulin") %>%
  ggplot(aes(x = study_week, y = acetate_mmol_kg, color = study_week)) + 
  geom_violin() + ggtitle("Acetate by Week") +
  xlab("Study week") + 
  ylab("Acetate (mmol/kg)") + 
  theme(legend.position = "none")

plot_but <- kghuang %>%
  filter(supplement_consumed == "inulin") %>%
  ggplot(aes(x = study_week, y = butyrate_mmol_kg, color = study_week)) + 
  geom_violin() + ggtitle("Butyrate by Week") +
  xlab("Study week") + 
  ylab("Butyrate (mmol/kg)") + 
  theme(legend.position = "none")

plot_pro <- kghuang %>%
  filter(supplement_consumed == "inulin") %>%
  ggplot(aes(x = study_week, y = propionate_mmol_kg, color = study_week)) + 
  geom_violin() + ggtitle("Propionate by Week") +
  xlab("Study week") + 
  ylab("Mean Propionate (mmol/kg)") + 
  theme(legend.position = "none")

plot_all <- plot_grid(plot_ace, plot_but, plot_pro)
plot_all
save_plot(filename = "figures/plot_all.pdf", plot = plot_all, nrow = 1, ncol = 2, base_aspect_ratio = 1.1)
```

Based on the data presented in the plot you generated above, identify which statistical test you would need to determine if there was a change in SCFA concentration between weeks 1 and 3 of the study. Check the assumptions appropriate for the test. [3 points]
```{r}
# assumptions
inulin_wk1 <- joined_df %>%
  filter(study_week == "week1", supplement_consumed == "inulin") %>%
  select(study_week, total_scfa)
shapiro.test(inulin_wk1$total_scfa) # Normality p-value = 4.106e-15
group_by(inulin_wk1) %>%
  summarize(sample_size = n()) # n = 432

inulin_wk3 <- joined_df %>%
  filter(study_week == "week3", supplement_consumed == "inulin")%>%
  select(study_week, total_scfa)
shapiro.test(inulin_wk3$total_scfa) # Normality p-value = 8.138e-8
group_by(inulin_wk3) %>%
  summarize(sample_size = n()) # n = 329

inulin <- joined_df %>%
  filter(supplement_consumed == "inulin") %>%
  select(study_week, total_scfa) 
bartlett.test(total_scfa ~ study_week, data = inulin) # Variance p-value = 0.9989 (both week1 and week3)

# test 
wilcox.test(x = inulin_wk1$total_scfa, y = inulin_wk3$total_scfa, alternative =  paired = FALSE) # p-value = 6.80e-6

# result?
# As the p-value from the Wilcoxon test was less than 0.05, there is a significant change in SCFA concentration from week 1 to week 3 in the participants who consumed Inulin. 
```

Based on the results of the test, conclude if there was a difference in SCFAs between weeks. 

<type answer here> 

### Question 10

Is there a correlation between the abundace of hydrogen generators and breath hydrogen? Create a plot that shows the relationship and conduct the appropriate statistical test(s). [2 points]
```{r}
# plot
plot_h2 <- all_data %>%
  filter(category == "hydrogen_gen") %>%
  ggplot(aes(x = fract, y = mean_h2)) + 
  geom_point() + ggtitle("Abundance of Hydrogen Generators vs. Breath Hydrogen") +
  xlab("Abundance of Hydrogen Generators") + 
  ylab("Mean Breath Hydrogen") + 
  theme(legend.position = "none")
plot_h2

save_plot(filename = "figures/plot_h2.pdf", plot = plot_h2, nrow = 1, ncol = 2, base_aspect_ratio = 1.1)

# test
m1 <- lm(fract ~ mean_h2, data = all_data)
summary(m1)

# result?
# p-value = 4.223e-5, adjusted R-squared value = 1.35e-4
```

What can you conclude from these results?

As the p-value is very small, there is significance, and thus correlation exists between abundance of hydrogen generators and breath hydrogen. However, the R-squared value is very small, and thus the quality of the linear fit is bad. 

Is there a correlation between the abundace of hydrogen consumers and breath methane? Create a plot that shows the relationship and conduct the appropriate statistical test(s). [2 points]
```{r}
# plot
plot_ch4 <- all_data %>%
  filter(category == "hydrogen_consumers") %>%
  ggplot(aes(x = fract, y = mean_ch4)) + 
  geom_point() + ggtitle("Abundance of Hydrogen Consumers vs. Breath Hydrogen") +
  xlab("Abundance of Hydrogen Consumers") + 
  ylab("Mean Breath Hydrogen") + 
  theme(legend.position = "none")
plot_ch4

save_plot(filename = "figures/plot_ch4.pdf", plot = plot_ch4, nrow = 1, ncol = 2, base_aspect_ratio = 1.1)

# test
m2 <- lm(fract ~ mean_ch4, data = all_data)
summary(m2)
# result?
# p-value = 2.2e-16, adjusted R-squared = 0.001751
```

What can you conclude from these results?

The result is similar to the previous example. As the p-value is very small, there is significance, and thus correlation exists between abundance of hydrogen consumers and breath methane. However, the R-squared value is very small, and thus the quality of the linear fit is bad. 

### Extra credit:

Is the abundace of hydrogen producers and consumers related? Create a plot, check assumptions, conduct the appropriate statistical test(s), and interpret the results. [1 point]
```{r}
# plot

# test(s)
```

<type answer here> 


-----
end